\documentclass[aspectratio=169]{beamer}

\usepackage[T1]{fontenc}
\usepackage{url}

\usetheme{Madrid}
\usecolortheme{dolphin}
\setbeamertemplate{navigation symbols}{}

\definecolor{proflogblue}{RGB}{36,82,128}
\definecolor{proofgreen}{RGB}{30,111,80}
\definecolor{limitorange}{RGB}{168,91,33}
\setbeamercolor{title}{fg=proflogblue}
\setbeamercolor{frametitle}{fg=proflogblue}
\setbeamercolor{block title}{bg=proflogblue,fg=white}
\setbeamercolor{block title alerted}{bg=limitorange,fg=white}
\setbeamercolor{block title example}{bg=proofgreen,fg=white}

\title[Executable SJAS in Proflog]{Executable Self-Justifying Axiom Systems in Proflog}
\subtitle{LOPSTR+PPDP 2026 System Description}
\author{James Torre}
\institute{\texttt{jpt4@proton.me}}
\date{LOPSTR+PPDP 2026}

\newcommand{\code}[1]{\texttt{#1}}
\newcommand{\isdbeta}{\(IS\#_D(\beta)\)}

\begin{document}

\begin{frame}
  \titlepage
  \vspace{-1.1em}
  \begin{center}
    \small System link: \url{https://github.com/autarkenterprises/proflog}
  \end{center}
\end{frame}

\begin{frame}{Core Claim}
  \begin{block}{System-description thesis}
    Willard-style self-justifying axiom systems can be made executable in
    Proflog by treating axiom bases, formula/proof codes, proof predicates,
    and generated self-consistency sentences as ordinary first-order objects.
  \end{block}

  \vspace{0.6em}
  \begin{itemize}
    \item Proflog supplies the proof-producing tableau kernel.
    \item SJAS supplies the weak arithmetic and restricted reflection target.
    \item The implementation exposes the programming-language boundary:
          reflection changes what the internal proof predicate may cite.
  \end{itemize}
\end{frame}

\begin{frame}{Why This Boundary Matters}
  \begin{columns}[T,onlytextwidth]
    \begin{column}{0.48\textwidth}
      \begin{block}{Arithmetic kept}
        \begin{itemize}
          \item constants \(0\), \(1\)
          \item successor-like binary numerals
          \item total addition
          \item enough coding to inspect formulas and proofs
        \end{itemize}
      \end{block}
    \end{column}
    \begin{column}{0.48\textwidth}
      \begin{alertblock}{Arithmetic withheld}
        \begin{itemize}
          \item no total multiplication symbol
          \item multiplication appears as \(\mathit{mult}(x,y,z)\)
          \item proof/code manipulation must respect that restriction
        \end{itemize}
      \end{alertblock}
    \end{column}
  \end{columns}

  \vspace{0.8em}
  The implementation makes the restriction operational rather than only
  metamathematical.
\end{frame}

\begin{frame}{Why Proflog Is a Useful Host}
  \begin{itemize}
    \item Proflog programs are first-order formulae with equality.
    \item Execution uses semantic tableau closure plus a Procedure Call rule.
    \item The same kernel can execute user programs and check encoded proof
          certificates.
  \end{itemize}

  \vspace{0.8em}
  \begin{exampleblock}{Key alignment}
    The deduction method \(D\) for the generated \isdbeta{} system is the
    ordinary Proflog first-order/equality tableau kernel.
  \end{exampleblock}
\end{frame}

\begin{frame}{Implemented Profile}
  \begin{columns}[T,onlytextwidth]
    \begin{column}{0.23\textwidth}
      \begin{block}{Inputs}
        \small
        profile choice\\
        finite \(\beta\)\\
        reflected clauses\\
        external clauses
      \end{block}
    \end{column}
    \begin{column}{0.24\textwidth}
      \begin{block}{Generated}
        \small
        Group-Zero/1\\
        Group-2 beta\\
        Group-2b reflected\\
        Group-3 SelfCons
      \end{block}
    \end{column}
    \begin{column}{0.25\textwidth}
      \begin{block}{Compiled}
        \small
        executable Proflog program\\
        \code{axiom-member}\\
        system/formula/proof codes
      \end{block}
    \end{column}
    \begin{column}{0.23\textwidth}
      \begin{block}{Queried}
        \small
        direct calls\\
        SJAS theorem calls\\
        certificate checks
      \end{block}
    \end{column}
  \end{columns}
\end{frame}

\begin{frame}[fragile]{Authoring Boundary}
  \begin{columns}[T,onlytextwidth]
    \begin{column}{0.48\textwidth}
      \begin{exampleblock}{Reflected clause}
\begin{verbatim}
demo(x) :- x = 1.
\end{verbatim}
        \small
        Executable by Proflog and included as a Group-2b axiom citeable by
        the internal proof predicate.
      \end{exampleblock}
    \end{column}
    \begin{column}{0.48\textwidth}
      \begin{alertblock}{External clause}
\begin{verbatim}
app(x) :- x = 1.
\end{verbatim}
        \small
        Executable by ordinary procedure call, but absent from the system code
        and generated self-consistency claim.
      \end{alertblock}
    \end{column}
  \end{columns}
\end{frame}

\begin{frame}[fragile]{Codes Are Inspectable}
  \begin{itemize}
    \item First prototype: finite labels for formulas and proof targets.
    \item Current compact form: byte/base-64 code terms.
    \item U-grounding form: public codes are binary numerals over the SJAS
          arithmetic vocabulary.
  \end{itemize}

  \vspace{0.7em}
\begin{verbatim}
(code-N b0 ... bN-1)

0, 1, (dbl n), (add n m)
\end{verbatim}

  \vspace{0.2em}
  The promoted relations consume codes; proof validity is not a host-side
  lookup table.
\end{frame}

\begin{frame}{Proof-Predicate Path}
  \begin{columns}[T,onlytextwidth]
    \begin{column}{0.32\textwidth}
      \begin{block}{Decode formula}
        \small
        \code{wff}\\
        formula classes\\
        \code{neg-pair}
      \end{block}
    \end{column}
    \begin{column}{0.32\textwidth}
      \begin{block}{Substitute}
        \small
        \code{subst-code/2}\\
        \code{subst-prf/4}\\
        Level-1 fixed point
      \end{block}
    \end{column}
    \begin{column}{0.32\textwidth}
      \begin{exampleblock}{Check proof}
        \small
        \code{tableau-proof/3}\\
        decode certificate\\
        invoke Proflog kernel
      \end{exampleblock}
    \end{column}
  \end{columns}

  \vspace{0.9em}
  A certificate succeeds only when the decoded proof closes the decoded target
  from the active generated system.
\end{frame}

\begin{frame}[fragile]{Example: Beta Axiom or Procedure}
  \begin{block}{Beta axiom}
\begin{verbatim}
forall x. mult(2, 2, x) -> composite(x)
\end{verbatim}
    \small
    \code{composite(4)} is an SJAS theorem, but there is no runtime
    \code{composite/1} procedure.
  \end{block}

  \begin{exampleblock}{Reflected procedure}
\begin{verbatim}
composite(x) :- mult(2, 2, x).
\end{verbatim}
    \small
    It is executable, citeable as Group-2b, and supports answer mode for
    \code{composite(x)}.
  \end{exampleblock}
\end{frame}

\begin{frame}{Evaluation Evidence}
  \begin{columns}[T,onlytextwidth]
    \begin{column}{0.48\textwidth}
      \begin{block}{Focused gate}
        \code{lein test-proflog-sjas}\\
        \vspace{0.3em}
        35 tests\\
        221 assertions\\
        0 failures, 0 errors
      \end{block}
    \end{column}
    \begin{column}{0.48\textwidth}
      \begin{block}{Slow SJAS selector}
        \code{lein test-proflog-sjas-slow}\\
        \vspace{0.3em}
        5 tests\\
        22 assertions\\
        0 failures, 0 errors
      \end{block}
    \end{column}
  \end{columns}

  \vspace{0.8em}
  Coverage includes reflected/external boundaries, U-grounding arithmetic,
  answer mode, partial synthesis, structural decoding, and invalid-certificate
  rejection.
\end{frame}

\begin{frame}{What This Does Not Claim}
  \begin{alertblock}{Limits of the current system}
    \begin{itemize}
      \item finite ordinary-tableau \isdbeta{} substrate, not every Willard
            variant;
      \item no Tab-1 theorem reuse or proof-list checker yet;
      \item open proof-code synthesis remains operationally expensive;
      \item consistency remains Willard's mathematical metatheorem, not a
            machine-checked proof in this implementation.
    \end{itemize}
  \end{alertblock}
\end{frame}

\begin{frame}{Submission Fit}
  \begin{itemize}
    \item Working system: public Proflog repository and focused examples.
    \item Novel aspect: executable SJAS profile inside a tableau logic
          programming system.
    \item Declarative-programming angle: reflection is an authoring choice that
          changes proof-citation power.
    \item Verification angle: object-language proof predicates route to real
          kernel certificate checks.
  \end{itemize}
\end{frame}

\begin{frame}{Takeaway}
  \begin{block}{Executable reflection with a visible cost model}
    Proflog makes the SJAS construction concrete enough to program against:
    generated finite axiom systems, arithmetized syntax, kernel-checked proof
    predicates, and a clear boundary between reflected and external code.
  \end{block}

  \vspace{0.8em}
  The next research step is to push more syntax construction and proof-code
  manipulation into the restricted U-grounding arithmetic so programs directly
  experience the absence of total multiplication.
\end{frame}

\end{document}
